Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2018-01-09 article/APM/article.tex
version submittet til APM 9/1-2018

\documentclass[a4paper,14pt]{article}
\usepackage[danish]{babel}
\usepackage{authblk}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathrsfs}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\usepackage{apalike}
\usepackage[utf8x]{inputenc}
\usepackage{nameref,hyperref}
% line numbers
%\usepackage[right]{lineno}
% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }
% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}
% array package and thick rules for tables
\usepackage{array}
% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

@@

\author[1]{Anne Helby Petersen}
\author[1]{Bo Markussen}
\author[3]{Karl Bang Christensen}
\affil[1]{Department of Public Health, University of Copenhagen, Copenhagen, Denmark}
\affil[2]{Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark}

@@

\begin{abstract}
Researchers often encounter data that is divided into distinct subsets. This might be due to multi-center sampling frames, or to variations in instruments, questionnaire item ordering or mode of administration. Often, one wishes to combine such data into one and conduct only a single analysis, but this is only meaningful if the subsets are similar in structure. Often, procedures for assessing whether such a data collapsing is allowed are ad hoc and require making full model assumptions. We present visual diagnostic tools for investigating differences in structure for two datasets containing different observations of the same variables. This approach, referred to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC), requires no directional nor hierarchical assumptions on the data and consists of three plots based on Principal Component Analysis (PCA), that address differences in the structures of the covariance matrices of the two datasets to yield an  understanding of where differences occur and why so. The functionality of the PCADSC tools is presented by use of data from the European Social Survey on psychological well-being in three European countries, namely Denmark, Sweden and Bulgaria. Here, we find that the concepts of psychological well-being are not the same when comparing Denmark and Bulgaria, and thus a comparison of e.g. mean scores in different well-being aspects is not meaningful. However, when comparing the two Scandinavian countries, Denmark and Sweden, we find very similar data structures and thus very similar concepts of well-being. Therefore, inter-country comparisons are warranted for these two countries.
\end{abstract}

%\keywords{Class file, \LaTeXe, \textit{SAGE Publications}}

@@

We propose three visual tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods uses principal component decomposition of the empirical covariance matrix in the two subsets to create intuitive visualizations of data structure differences. The proposed tools are implemented in the statistical software \texttt{R} in our package, \texttt{PCADSC} \cite{PCADSC}.

@@

PCADSC compares the covariance matrices of the two subsets of a dataset. If all variables in the datasets are jointly normal with known means, these are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing interrelations between variables. This makes the empirical covariance matrix a reasonable place to start looking for differences in data structures. Computing empirical covariance matrices for the two subsets and comparing these entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets.

@@

While we present the three plot types from a purely theoretical point of view below, we also provide example illustrations as a supplement to the somewhat technical definitions. These plots are available in Figs~\ref{plot.simCE}, \ref{plot.simAngle} and \ref{plot.simChroma} for the CE-, angle- and chroma plots, respectively, and they are based on two simulated datasets:
\begin{description}
\item[Dataset A:] This dataset contains 1000 independent simulations from the same underlying model, namely \textit{model 1} from Fig~\ref{fig.simGraphs}. The observations are randomly divided into two groups.
\item[Dataset B:] This dataset contains 500 independent simulations from \textit{model 1} from Fig~\ref{fig.simGraphs} and 500 independent simulations from \textit{model 2} in the same figure. The observations are naturally divided into groups corresponding to which model they were simulated from.
\end{description}
Further details about the data simulations are available in the \nameref{appendix.simData}. In dataset A, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures. In dataset B, on the other hand, there are two fundamentally different, underlying datastructures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.

\begin{figure}[!h]
\caption{{\bf Graphical representations of the two models used for simulating example data for the PCADSC plots.} Square nodes represent observed variables, while rounded nodes represent latent variables.  Arrows are used to illustrate causal links, and thus we e.g. find in \textit{Model 1} that the observed variables $v_3$ and $v_4$ are caused by a common, unobserved variable, namely $X_1$. Further details on the models and the simulated data are available in the \nameref{appendix.simData}, including covariance matrices for the two models.}
\label{fig.simGraphs}
\end{figure}

\subsubsection*{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of variance components, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
&(0,0), \;
(\lambda_1,\lambda_{11}-\lambda_{21}), \;
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), \;
\ldots, \\
&\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
\end{align*}
Note that due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are. Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does. Two examples of cumulative eigenvalue plots are available in Fig~\ref{plot.simCE}.

\begin{figure}[!h]
\caption{{\bf CE plots for the simulated datasets.} Dataset A, which includes no differences in data structures, is found in the top panel, and dataset B, where the observations stem from two different underlying models, is presented in the bottom panel. Both plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures. The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures. In the top panel, we see a cumulative eigenvalue curve that falls well within the confidence region, thereby suggesting no difference in eigenvalues (as expected). This also agrees with the large $p$-values and the fact that only a single underlying model was used for generating this data. In the bottom panel, the conclusion is reversed, which is also in accordance with the nature of the simulated data, which contains two distinct groups.}
\label{plot.simCE}
\end{figure}

In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
\begin{align*}
\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} {\bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)}^2.
\end{align*}
We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_1$ and $n_2$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics. This should be done a large (e.g. 10000) number of times. Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to %even larger
test statistics at least as large as %than
the one we found for the original datasets.

@@

\subsubsection*{The angle plot}
This plot simultaneously compares the variation components and the loadings. Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets. Then the empirical correlation matrix for the first dataset, $S_1$, has the following orthogonal decomposition in the coordinate system of the second dataset
\begin{equation*}
S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)}^\top,
\end{equation*}
and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k}^\top \eta_{2j} \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j}^\top \eta_{1k} \rvert, &
\theta_{jk} &= \acos(\lvert \eta_{1k}^\top \eta_{2j} \rvert).
\end{align*}
Note that for two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the sample correlation. Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance component contributions. The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length. Fig~\ref{plot.simAngle} presents two examples of angle plots based on simulated data.

\begin{figure}[!h]
\caption{{\bf Angle plots for the simulated datasets.}Dataset A is used in the top panel and dataset B is used in the bottom panel. The blue show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse. For dataset A, we find very short off-diagonal arrows, suggesting that the two PCA decompositions agree on the relative placement of the information in  the data among the 6 PCs.  For dataset B, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}
\label{plot.simAngle}
\end{figure}

@@

\subsubsection*{The chroma plot}
This plot compares the loadings of the two datasets. The chroma plot consists of two panels, one for each dataset, made up of colored bars. These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions. More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors. The width of the $j$th colored segment is given by
\begin{align*}
\omega_{1ij} &= \frac{\lvert\eta_{1ij}\rvert}{\sum_{k=1}^d \lvert\eta_{1ik}\rvert}, &
\omega_{2ij} &= \frac{\lvert\eta_{2ij}\rvert}{\sum_{k=1}^d \lvert\eta_{2ik}\rvert},
\end{align*}
where $\eta_{1ij}$ and $\eta_{2ij}$ denotes the $j$th entry of $\eta_{1i}$ and $\eta_{2i}$, respectively. Due to the indeterminacy of the sign, all the signs are removed from the coefficients in the loadings. The bars are ordered according to the variation components and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance component contributions,
\begin{align*}
\tilde\sigma^c_{1i} &= \frac{\sum_{j=1}^i \lambda_{1j}}{\sum_{k=1}^d \lambda_{1k}} = \frac{\sum_{j=1}^i \lambda_{1j}}{d}, &
\tilde\sigma^c_{2i} &= \frac{\sum_{j=1}^i \lambda_{2j}}{\sum_{k=1}^d \lambda_{2k}} = \frac{\sum_{j=1}^i \lambda_{2j}}{d}.
\end{align*}
Especially when $d$ is large, we recommend plotting only a select set of interesting principal components, e.g. as identified by use of the angle plot. In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{1i} = \frac{\lambda_{1i}}{d}$ and $\tilde\sigma_{2i} = \frac{\lambda_{2i}}{d}$. In Fig~ \ref{plot.simChroma}, two examples of chroma plots are available.

\begin{figure}[!h]
\caption{Chroma plots for the simulated datasets, with dataset A in the top and dataset B in the bottom. Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance ($\tilde\sigma_{ij}^c$). For dataset A, we see very similar visual patterns across the two panels, corresponding well with the fact that both groups are drawn from the same underlying model. For dataset B, we see different color compositions for the bars in the left- and the right panel, illustrating that the data structures are not identical for the two groups.}
\label{plot.simChroma}
\end{figure}

The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions. For each component, the loadings describe how influential the different variables are on that component. Therefore, the chroma plot allows us to make qualitative statements about the original datasets, such as \textit{"variable $x$ is generally more influential in subset 1 than it is in subset 2"}, thereby helping us to understand where and why potential data structure differences are found.

@@

% But the problem is beyond translation efforts: The main question really is whether or not there exist such a thing as a universal, internationally valid concept of happiness. If not, the very idea of translation is meaningless, since different aspects of psychological well-being or happiness simply do not have the same relative meaning in different cultural and socio-economic settings. This is in fact a question concerning comparability of data structures.

@@

All computations and figures presented in this section were created using our \R{R} package \R{PCADSC} \cite{PCADSC}.

@@

The ESS 2012 data contain a total of 626 variables collected from 54673 citizens of 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}. More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in the \nameref{table:items}. We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This corresponds to approximately 9\% of the observations in the Danish sample, 20\% in the Bulgarian sample and only 6\% in the Swedish sample. All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations. Table~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.

\begin{table}[!ht]
\centering
\begin{tabular}{lccc}
  \hline
  & Denmark & Bulgaria & Sweden \\
  \hline
 Evaluative wellbeing & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
 Emotional wellbeing & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
Functioning & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
Vitality & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
 Community wellbeing & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
Supportive relationships & 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
   \hline
\end{tabular}
\caption{Median values of each of the six dimensions of psychological well-being, stratified by country. 25 and 75 percentiles are listed in parentheses. Note that the scales are constructed such that they all run from 0-10.}
\label{tableDistr}
\end{table}

>>>>

\documentclass[a4paper,14pt]{article}
\usepackage[danish]{babel}
\usepackage{authblk}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathrsfs}
%\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\usepackage{apalike}
\usepackage[utf8x]{inputenc}
%\usepackage{nameref,hyperref}
% line numbers
%\usepackage[right]{lineno}
% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }
% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}
% array package and thick rules for tables
\usepackage{array}
% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

@@

\author[1]{}
%\author[1]{Anne Helby Petersen}
%\author[1]{Bo Markussen}
%\author[3]{Karl Bang Christensen}
%\affil[1]{Department of Public Health, University of Copenhagen, Copenhagen, Denmark}
%\affil[2]{Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark}

@@

\renewcommand{\abstractname}{Abstract}
\renewcommand{\figurename}{Figure}
\renewcommand{\tablename}{Table}

\begin{abstract}
Researchers often encounter data that is divided into distinct subsets. This might be due to multi-center sampling frames, or to variations in instruments, questionnaire item ordering or mode of administration. Often, one wishes to combine such data into one and conduct only a single analysis, but this is only meaningful if the subsets are similar in structure. Common procedures for assessing whether such a data collapsing is allowed are ad hoc and require making full model assumptions. We present visual diagnostic tools for investigating differences in structure for two datasets containing different observations of the same variables. This approach, referred to as \textit{Principal Component Analysis-based Data Structure Comparisons} (PCADSC), requires no directional nor hierarchical assumptions on the data and consists of three plots based on Principal Component Analysis (PCA), that address differences in the structures of the covariance matrices of the two datasets to yield an  understanding of where differences occur and why so. The functionality of the PCADSC tools is presented by use of data from the European Social Survey on psychological well-being in three European countries, namely Denmark, Sweden and Bulgaria. Here, we find that the concepts of psychological well-being are not the same when comparing Denmark and Bulgaria, and thus a comparison of e.g. mean scores in different well-being aspects is not meaningful. However, when comparing the two Scandinavian countries, Denmark and Sweden, we find very similar data structures and thus very similar concepts of well-being. Therefore, inter-country comparisons are warranted for these two countries.
\end{abstract}

@@

We propose three visual tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods use principal component decomposition of the empirical covariance matrix in the two sub sets to create intuitive visualizations of data structure differences. The proposed tools are implemented in the \texttt{R} package \texttt{PCADSC} \cite{PCADSC}.

@@

PCADSC compares the covariance matrices of the two subsets of a dataset. If all variables in the datasets are jointly normal with known means, the covariance matrices are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing interrelations between variables. This makes the empirical covariance matrix a reasonable place to start looking for differences in data structures. Computing empirical covariance matrices for the two subsets and comparing these entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets.

@@

While we present the three plot types from a purely theoretical point of view below, we also provide example illustrations as a supplement to the somewhat technical definitions. These plots are available in Figures \ref{plot.simCE}, \ref{plot.simAngle} and \ref{plot.simChroma} for the CE-, angle- and chroma plots, respectively, and they are based on two simulated datasets:
\begin{description}
\item[Dataset A:] This dataset contains 1000 independent simulations from the same underlying model, namely \textit{model 1} from Figure~\ref{fig:simGraphs}. The observations are randomly divided into two groups.
\item[Dataset B:] This dataset contains 500 independent simulations from \textit{model 1} from Figure~\ref{fig:simGraphs} and 500 independent simulations from \textit{model 2} in the same figure. The observations are naturally divided into groups corresponding to which model they were simulated from.
\end{description}

In \textit{Model 1} the observed variables $v_3$ and $v_4$ are caused by a common, unobserved variable $X_1$. Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models. Further details about data simulation is also available in the supplementary material. In dataset A, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures. In dataset B, on the other hand, there are two fundamentally different, underlying data structures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.

\begin{figure}[!h]
\caption{[FIGURE 1 HERE]}
% Graphical representations of models used for simulating example data. Square nodes represent observed variables, while rounded nodes represent latent variables. Arrows are used to illustrate causal links.
\label{fig:simGraphs}
\end{figure}

\subsubsection*{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the correlation matrix. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of variance components, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
&(0,0), \;
(\lambda_1,\lambda_{11}-\lambda_{21}), \;
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), \;
\ldots, \\
&\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg)
\end{align*}
Note that due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are. Moreover, a positive cumulative difference implies that dataset 1 holds more information in the first components than dataset 2 does. Two examples of cumulative eigenvalue plots are available in Figure \ref{fig:plot.simCE}.

\begin{figure}[!h]
\caption{[FIGURE 2 HERE]}
\label{plot.simCE}
% CE plots for the simulated datasets. Dataset A (with no differences in data structure) in top panel, and dataset B (where observations stem from different models) in bottom panel. Plots are annotated with the $p$-values of the Kolmogorov-Smirnov (KS) and the Cram\'er-von Mises (CvM) tests of the null hypothesis of no difference in data structures. The bold, black lines illustrate the observed cumulative eigenvalue differences, while the shaded areas are 95 \% confidence bands under the null hypothesis of no difference in data structures.
\end{figure}

In the top panel, we see a cumulative eigenvalue curve that falls well within the confidence region, thereby suggesting no difference in eigenvalues (as expected). This also agrees with the large $p$-values and the fact that only a single underlying model was used for generating this data. In the bottom panel, the conclusion is reversed, which is also in accordance with the nature of the simulated data, which contains two distinct groups. In order to test whether these cumulative differences are statistical artefacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
\begin{align*}
\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} {\bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)}^2.
\end{align*}
We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and separately standardized datasets into two new datasets of $n_1$ and $n_2$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics. This should be done a large (e.g. 10000) number of times. Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to %even larger
test statistics at least as large as %than
the one we found for the original datasets.

@@

\subsubsection*{The angle plot}
This plot simultaneously compares the variation components and the loadings. Let $\lambda_{\max} = \max\{ \lambda_{11}, \lambda_{21} \}$ be the largest variation component for the two datasets. Then the empirical correlation matrix for the first dataset, $S_1$, has the following orthogonal decomposition in the coordinate system of the second dataset
\begin{equation*}
S_1 = \sum_{k=1}^d \lambda_{1k} \eta_{1k} \eta_{1k}^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)
{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \eta_{2j} (\eta_{2j}^\top \eta_{1k}) \Bigg)}^\top,
\end{equation*}
and we have a similar decomposition of $S_2$ in the coordinate system of the first dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$ and the angle $\theta_{kj}$ are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{1k}}{\lambda_{\max}}} \lvert \eta_{1k}^\top \eta_{2j} \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{2j}}{\lambda_{\max}}} \lvert \eta_{2j}^\top \eta_{1k} \rvert, &
\theta_{jk} &= \acos(\lvert \eta_{1k}^\top \eta_{2j} \rvert).
\end{align*}
Note that for two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the sample correlation. Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance component contributions. The absolute value of the projection $\eta_{1k} \eta_{2j}^\top$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_1$ and $S_2$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length. Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data.

\begin{figure}[!h]
\caption{[FIGURE 3 HERE]}
% Angle plots for the simulated datasets. Dataset A in top panel and dataset B in bottom panel. Blue arrows show the principal components of the observations in group 1 decomposed in the coordinate system of the principal components of group 2, while the red arrows illustrate the reverse.
\label{plot.simAngle}
\end{figure}

For dataset A, we find very short off-diagonal arrows, suggesting that the two PCA decompositions agree on the relative placement of the information in  the data among the 6 PCs.  For dataset B, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.

@@

\subsubsection*{The chroma plot}
This plot compares the loadings of the two datasets. The chroma plot consists of two panels, one for each dataset, made up of colored bars. These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions. More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors. The width of the $j$th colored segment is given by
\begin{align*}
\omega_{1ij} &= \frac{\lvert\eta_{1ij}\rvert}{\sum_{k=1}^d \lvert\eta_{1ik}\rvert}, &
\omega_{2ij} &= \frac{\lvert\eta_{2ij}\rvert}{\sum_{k=1}^d \lvert\eta_{2ik}\rvert},
\end{align*}
where $\eta_{1ij}$ and $\eta_{2ij}$ denotes the $j$th entry of $\eta_{1i}$ and $\eta_{2i}$, respectively. Due to the indeterminacy of the sign, all the signs are removed from the coefficients in the loadings. The bars are ordered according to the variation components and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance component contributions,
\begin{align*}
\tilde\sigma^c_{1i} &= \frac{\sum_{j=1}^i \lambda_{1j}}{\sum_{k=1}^d \lambda_{1k}} = \frac{\sum_{j=1}^i \lambda_{1j}}{d}, &
\tilde\sigma^c_{2i} &= \frac{\sum_{j=1}^i \lambda_{2j}}{\sum_{k=1}^d \lambda_{2k}} = \frac{\sum_{j=1}^i \lambda_{2j}}{d}.
\end{align*}
Especially when $d$ is large, we recommend plotting only a select set of interesting principal components, e.g. as identified by use of the angle plot. In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{1i} = \frac{\lambda_{1i}}{d}$ and $\tilde\sigma_{2i} = \frac{\lambda_{2i}}{d}$. In Figure  \ref{plot.simChroma}, two examples of chroma plots are available.

\begin{figure}[!h]
\caption{[FIGURE 4 HERE]}
% Chroma plots for the simulated datasets. Dataset A in top panel and dataset B in bottom panel. Each bar represents a principal component and the bars are annotated with their cumulative percentages of explained variance.
\label{plot.simChroma}
\end{figure}

For dataset A, we see very similar visual patterns across the two panels, corresponding well with the fact that both groups are drawn from the same underlying model. For dataset B, we see different color compositions for the bars in the left- and the right panel, illustrating that the data structures are not identical for the two groups. The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions. For each component, the loadings describe how influential the different variables are on that component. Therefore, the chroma plot allows us to make qualitative statements about the original datasets, such as \textit{"variable $x$ is generally more influential in subset 1 than it is in subset 2"}, thereby helping us to understand where and why potential data structure differences are found.

@@

@@

All computations and figures presented in this section were created using the \R{R} package \R{PCADSC} \cite{PCADSC}.

@@

The ESS 2012 data contain a total of 626 variables collected from 54673 citizens of 29 countries. Here, we will only work with a subset of 35 questionnaire items that are all related to psychological well-being. These 35 items can be divided into 6 distinct scales, namely \textit{Evaluative wellbeing}, \textit{Emotional wellbeing}, \textit{Functioning}, \textit{Vitality}, \textit{Community wellbeing} and \textit{Supportive relationships}. More details on these scales can be found in \cite{ESStopline5} and the relationship between questionnaire items and scales is summarized in the supplementray material. We represent each of the scales by a single variable, which is calculated as the average score within the items related to that variable and scaled such that it takes a value between 0 and 10. For simplicity, we use only complete cases for this construction and thus exclude all participants that did not answer all the 35 questionnaire items used below. This corresponds to approximately 9\% of the observations in the Danish sample, 20\% in the Bulgarian sample and only 6\% in the Swedish sample. All in all, we have $n_{DK} = 1498$ complete case observations in the Danish sample, $n_{BG} = 1798$ observations in the Bulgarian sample and $n_{SE} = 1736$ Swedish observations. Table~\ref{tableDistr} summarizes the marginal distributions of the six dimensions of psychological well-being, stratified by country.

\begin{table}[!ht]
\centering
\begin{tabular}{lccc}
\hline
  & Denmark & Bulgaria & Sweden \\
\hline
Evaluative wellbeing    & 8.75 (8.00, 9.50) & 5.00 (3.50, 7.00) & 8.00 (7.00, 9.00) \\
Emotional wellbeing     & 8.33 (7.22, 8.89) & 6.67 (5.00, 7.78) & 7.78 (6.67, 8.89) \\
Functioning             & 7.57 (6.93, 8.21) & 6.68 (5.50, 7.68) & 7.04 (6.39, 7.68) \\
Vitality                & 7.50 (6.67, 8.33) & 7.50 (5.83, 8.33) & 7.50 (6.67, 9.17) \\
Community wellbeing     & 6.77 (5.83 ,7.57) & 4.67 (3.70, 5.70) & 6.57 (5.66, 7.37) \\
Supportive relationships& 8.25 (7.42, 8.92) & 7.25 (6.17, 8.08) & 8.25 (7.42, 8.75) \\
\hline
\end{tabular}
\caption{Median values of each of the six dimensions of psychological well-being, stratified by country. 25 and 75 percentiles are listed in parentheses. Note that the scales are constructed such that they all run from 0-10.}
\label{tableDistr}
\end{table}

\subsection*{Comparing Denmark and Bulgaria}
Figure \ref{plotBG.cehair} presents the CE plot and the angle plot obtained from comparing the Danish and Bulgarian psychological well-being scales. The CE plot show a remarkable degree of lacking comparability: The cumulative differences in the eigenvalues by far exceed what could come about randomly if there really were no difference in the data structures. This is also confirmed by the Kolmogorov-Smirnov and the Cram\'er-von Mises tests, which both result in $p$-values that are virtually zero.
